EuPPollNet: A European database of plant-pollinator networks

### 2024-09-21 Manuscript/EuPPollNet_word.qmd
Affiliations update

@@

**Main Types of Variables Included:** EuPPollNet contains `r prettyNum(interaction_number, big.mark=",")` interactions between plants and pollinators from `r prettyNum(n_networks, big.mark=",")` distinct networks (i.e., distinct sampling event in space or time), which belong to `r study_number` different studies distributed across `r country_number` European countries. In addition, information about sampling methodology, habitat type, bio-climatic region, and further taxonomic rank information for both plant and pollinator species are also provided (i.e., family, order and phylum).

**Spatial location and grain:** The database contains `r prettyNum(n_sites, big.mark=",")` different sampling locations from natural and anthropogenic habitats that fall in `r n_bioregion` different bio-climatic regions. All records are geo-referenced and presented in the World Geodetic System 1984 (WGS84).

@@

The interactions between different plant and pollinator species within a community form complex networks. Macro-ecological analyses of the topology of these networks have revealed common properties, such as truncated power-law degree distributions [@jordano1987] or nestedness [@bascompte2003]. Large-scale analyses across multiple studies can quantify patterns across geographic regions [@olesen2002; @traveset2016] or environmental gradients [@ramos2010; @rech2016; @saunders2023] that cannot be examined in a single study. Although macro-ecological approaches that use ecological interactions make significant contributions to knowledge [@windsor2023], such synthesis work must consider variation across studies in the spatio-temporal nature of the data [@burkle2011; @trojelsgaard2013]. For instance, plant-pollinator studies tend to differ in sampling effort and methodology which affect the structure of the resulting plant-pollinator networks [@gibson2011; @jordano2016; @schwarz2020]. Most plant-pollinator networks have unobserved interactions [@olesen2011; @chacoff2012], and thus research that attempts to synthesize across published studies must have access to raw data on interactions in order to statistically account for sampling effort and completeness. This emphasizes the importance of providing data in its rawest possible form in datasets that will be utilized for synthesis and macro-ecological studies.

Europe is one of the continents with a larger amount of available biodiversity data [@proencca2017], yet still exhibits major gaps [@wetzel2018; @bennett2018]. While species checklists need to be treated carefully, especially at a macro-ecological scale [@grenie2023], the growing number of European plant and pollinator checklists [@reverte2023], along with occurrence data [@zattara2021], is setting a foundation for the conservation of its flora and their pollinators. However, species richness is just one component of biodiversity and documenting the interaction between plants and pollinators is essential for understanding biodiversity change [@jordano2016]. Numerous works have studied plant-pollinator interactions in the last decades, originating thousands of plant-pollinator interaction networks worldwide. Several initiatives have tried to integrate plant-pollinator interaction data into databases such as *Mangal* [@poisot2016] or *GloBI* [@poelen2014], resulting in numerous large scale comparative analyses that have enhanced our understanding of the ecology of plants and pollinators [e.g., European wild bee data trends\; @marshall2024]. Despite all these resources, Europe lacks accessible harmonized plant-pollinator interaction data that allow researchers to evaluate plant-pollinator interactions at a European level, which will guide research efforts, conservation planning and will set a foundation for future global change research. For example, only over a dozen of European plant-pollinator networks are included in *Mangal*, while *GloBI* focuses on pairwise interactions disconnected from the community context. Assembling and curating the existing information on EU plant-pollinator networks will guide research efforts, conservation planning and will set a foundation for future global change research.

Here, we present the European Plant-Pollinator Networks database (EuPPollNet), which contains harmonized information on plant-pollinator interactions at the European level. The pollinator taxonomic groups include the main orders of entomofauna that visit and pollinate flowering plants in Europe. These comprise insect species from the orders Hymenoptera, Diptera, Lepidoptera, and Coleoptera [@willemstein1987; @potts2015], accounting for almost the totality of recorded interactions in EuPPollNet (`r format(round(main_orders_percent, 2), nsmall = 2)`%). To understand the scope of the database, we examined the taxonomic and sampling coverage of the different plant and pollinator species at the European level with the help of the most up-to-date species checklists and rarefaction analyses. The overall number of potential pollinators and flowering plants in Europe was estimated with the help of the different checklists and sampling coverage. Finally, for bees and plants, we evaluated if there is a phylogenetic signal in the presence-absence of interaction data.

We investigated how two key structural metrics of plant-pollinator networks, connectance and nestedness, change across latitudes and bioclimatic regions. Networks that are more nested than expected by chance exhibit greater redundancy in their ecological functions and more stability in the face of species loss [@thebault2010]. EuPPollNet contains one of the largest sets of plant-pollinator networks collated to date, providing a unique opportunity to examine the prevalence of nestedness, which has only been evaluated with a relatively small number of networks [@bascompte2003; @staniczenko2013; @payrato2019]. EuPPollNet aims to cover a wide range of taxonomic groups and habitats, while also providing other variables of interest that define the ecological context and sampling methods of the study. In addition, EuPPollNet offers a transparent and accessible workflow of its data management and species harmonization that allows the database to be reused and to expand over time. This database provides a large number of community-level networks with curated and harmonized data, distinguishing it from other currently available resources that contain plant-pollinator interactions. We expect that EuPPollNet can be used to evaluate macro-ecological processes in plant-pollinator networks, guide conservation planning, and set a baseline for global change research.

@@

The EuPPollNet database includes published and unpublished studies compiled initially by a wide number of researchers and institutions within the European continent. As this database is the result of one of the working packages of the European project Safeguard, first, data was directly asked from members of the Safeguard project. Second, the request was extended to data owners outside of the project. These other data owners were identified by direct communication with colleagues suggested by Safeguard members and by directly searching for studies on Google Scholar of under-represented regions within the database. To maintain high quality standards that will support robust future ecological research, we only included studies meeting the following criteria: 1) studies that contained time- and geo-referenced records of plant-pollinator interactions; and 2) studies with phyto-centric plant-pollinator networks with quantitative visitation data.

@@

The database contains `r study_number` independent published and unpublished studies conducted during the time period `r min_year` - `r max_year` in `r country_number` different countries (**Figure 1a** and **Figure 1b**). The studies differ in sampling effort and methodology, and thus documenting sampling methods and sampling effort is an important feature of EuPPollNet. Most studies took place within a single flowering season (`r format(round(one_year, 2), nsmall = 2)`%), sampled a given location an average of `r format(round(average_sampling_days, 2), nsmall=2)` days, and documented interactions mostly by using `r  most_common_method[[1]]` as the sampling method (`r format(round(most_common_method[[3]], 2), nsmall = 2)`%). The database includes a total of `r prettyNum(interaction_number, big.mark=",")` distinct interactions, considering interaction as the contact of a pollinator individual to the reproductive structure of a particular plant. Most of the pollinator species in the database belong to the orders Hymenoptera and Diptera, each comprising approximately 1,000 species in the database. However, the majority of plant-pollinator interactions are from Hymenoptera species (`r format(round(Hymenoptera_interactions, 2), nsmall = 2)`%; **Figure 1c**). Notably,  _Apis mellifera_ represents `r format(round(apis_percent, 2), nsmall = 2)`% of the total interaction records from the database and an average of `r format(round(apis_average_network, 2), nsmall = 2)`% of the total interactions per network.

```{r echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, out.width="110%", fig.cap= "\\textbf{Figure 1}. (\\textbf{a}) Locations of the studies in EuPPollNet showing the total number of pollinator (i.e., orange heptagon) and plant (i.e., green circles) species per study. The sizes of these shapes are proportional to the respective species counts. For visualization purposes, we have focused only on the European region with studies and selected a single location per study. (\\textbf{b}) Number of studies by year in EuPPollNet. (\\textbf{c}) Number of species and interactions across the four main pollinator orders in EuPPollNet."}

@@

p2  = ggplot(data2, aes(Year, Counts)) +
geom_col(fill = "gray25", width = 0.9, color="black", linewidth = 0.4) +
theme_light() +
theme(plot.margin=unit(c(0,0,0,0), "mm"),
  axis.title.y = element_text(face = "bold", size = 7, vjust = -2),
  axis.title.x = element_text(face = "bold", size = 8),
  axis.text.y = element_text(size = 6),
  axis.text.x = element_text(size = 6,  angle = 90), 
  strip.text.x = element_text(size=12),
   axis.ticks.x = element_line(color="black"),
  strip.background = element_rect(
  color="black", fill=NA, size=0.5, linetype="solid")) + 
ylab("Number of studies") +
coord_cartesian(expand = F, clip = "off") + 
theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  panel.background = element_blank(),
  panel.border = element_rect(colour = "black", fill=NA, size=1)) +
theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(),
panel.border = element_blank(),
axis.line = element_line(color = 'black')) +
scale_y_continuous(breaks = c(0, 4,8,12), labels = c("0", "4","8", "12"), limits = c(0,12)) +
scale_x_continuous(breaks = seq(min(data2$Year), max(data2$Year), by = 2), limits = c(min(data2$Year),max(data2$Year)))+
ggtitle("(b)") +
theme(plot.title = element_text(size = 10,hjust = -0.3, vjust = -5))

main_orders = c("Hymenoptera", "Diptera", "Lepidoptera", "Coleoptera")
#Select main orders
data1 = data %>% 
select(Pollinator_order, Pollinator_accepted_name) %>% 
filter(Pollinator_order %in% main_orders) %>% 
group_by(Pollinator_order) %>% 
summarise(Species = n_distinct(Pollinator_accepted_name)/n_distinct(.),
          Interactions = length(Pollinator_accepted_name)/ nrow(.))

# Combine Interactions and Species into a single column
data_long = pivot_longer(data1, cols = c(Interactions, Species), names_to = "Variable", values_to = "Value")

@@

#Plot
p3 = ggplot(data_long, aes(x = Variable, y = Value, fill = Pollinator_order)) + 
geom_bar(stat = "identity", width = 0.7) +
theme_classic() +
scale_fill_viridis(discrete=TRUE, direction = -1, name="Pollinator order") +
xlab(NULL) +
ylab("Percentage") + 
scale_y_continuous(labels = percent_format(), expand = c(0, 0)) +
theme(axis.text = element_text(size=6), 
     axis.title.y = element_text(face = "bold", size = 7, vjust = -2),  
     axis.title.x = element_text(face="bold", size = 7),
     legend.position = "top",
     legend.title = element_text(size = 5, face = "bold"),
     legend.text = element_text(size = 5),
     legend.key.size = unit(3, "mm"),
     legend.margin=margin(0,0,0,0),
     legend.box.margin=margin(-8,-5,-5,-5)) +
guides(fill=guide_legend(nrow=2,byrow=TRUE))+
ggtitle("(c)") +
theme(plot.title = element_text(size = 10,hjust = -0.3, vjust = -6))

@@

The EuPPollNet database is available in both .csv and .rds formats. The files contain a total of `r nrow(my_tibble) + 1` columns that depict where and when the plant-pollinator interaction was observed (**Table 1**). Each row depicts a single interaction between a plant and a pollinator species. To build a plant-pollinator network matrix within a single flowering season at a site level, users only need to group interactions by plant and pollinator species, site, study and year. Information about habitat type and bioclimatic region are also provided in this file. In addition, the interaction dataset includes a column that describes the presence or absence of floral counts for each study. The flower count data is provided in a separate file (.csv or .rds) that can be merged with the interaction data through the "Flower_data_merger" column. Two thirds of studies contain information on floral abundance, but the methods and units differ greatly across studies. Finally, metadata at the study level is provided, including information about the authors, digital object identifier if available, sampling time and taxonomic coverage of the main pollinator groups for each study.

@@

To assess the completeness of plant and pollinator species in the EuPPollNet database at the European level, we used the aforementioned checklists for plants and pollinators. Specifically for plants, we refined the checklist to include only European flowering plants and excluded taxonomic groups not associated with biotic pollination. We did this by first excluding the families considered to have exclusively a wind pollination mode [see @culley2002], and then by filtering out the genera with wind or non-biotic pollination from families that exhibit both biotic and non-biotic pollination modes. Additionally, we manually included some exotic species and added unresolved species names that were not present in the accepted names of the checklist at the current version of usage. For pollinators, we compared only the taxonomic coverage of bees, syrphids and butterflies by using their species checklists at the European level. The number of pollinator species at the European level was estimated by first estimating coverage for the three taxa that have checklists (bees, syrphids and butterflies). The mean coverage across these three groups (mean coverage = `r round(mean_coverage, 2)`; sd = `r round(sd_coverage,2)`) was then used for the other taxa for which there is no checklist to extrapolate the number of pollinating species in Europe.

@@

To provide a general overview of the structure of plant-pollinator networks in EuPPollNet, we quantified connectance and nestedness for each network, and examined how these network metrics change across latitude of studies and bioclimatic regions in Europe. We selected these two network metrics as they are commonly evaluated in plant-pollinator network studies and capture structural properties with a straightforward interpretation. We implemented “standardised” versions of connectance and nestedness to account for the effect of sampling effort on network metrics. As connectance is negatively associated with network size [@jordano1987], we evaluated how network connectance was associated with the number of species (i.e., log of geometric mean of plants and pollinators) and extracted the residuals from this association (i.e., residual connectance) as a measurement of corrected connectance. The relationship between residual connectance and species richness was investigated using a beta-regression implemented with the package **betareg** [@cribari2010]. We used NODFc to compare nestedness across networks, as it corrects by connectance and the number of species [@song2017]. This metric was calculated using the **maxnodf** package [@hoeppke2021]. Both residual connectance and NODFc were used as dependent variables to evaluate their association with latitude. In addition, we quantified the association between connectance and nestedness with the number of species per network using a Kendall rank correlation coefficient to compare the strength of associations between network structures and species richness across both network metrics.

>>>>

#Check number of species from Hymenop, Dipte, Lepidop and Coleop

pollinator_spp = data %>% 
select(Pollinator_rank, Pollinator_accepted_name, Interaction) %>% 
filter(Pollinator_rank == "SPECIES") %>% 
distinct(Pollinator_accepted_name) %>% 
nrow(.)

pollinator_spp_main_orders = data %>% 
select(Pollinator_order, Pollinator_rank, Pollinator_accepted_name, Interaction) %>% 
filter(Pollinator_rank == "SPECIES") %>% 
filter(Pollinator_order == "Hymenoptera"|
       Pollinator_order == "Diptera"|
       Pollinator_order == "Lepidoptera"|
       Pollinator_order == "HymenoColeopteraptera") %>% 
distinct(Pollinator_accepted_name) %>% 
nrow(.)

spp_percent_main_order = pollinator_spp_main_orders/pollinator_spp *100

#Calculate percentage of studies that recorded the main 4 orders
metadata = readRDS("../Data/3_Final_data/Metadata.rds")
#Some studies are divided by year
#Get an average of the pollinator percentages across years first
#Fix NA's
metadata_study = metadata %>%
group_by(Study_id) %>% 
mutate(Hymenoptera_percent = as.numeric(na_if(Hymenoptera_percent, "NA"))) %>%  
mutate(Diptera_percent = as.numeric(na_if(Diptera_percent, "NA"))) %>%  
mutate(Lepidoptera_percent = as.numeric(na_if(Lepidoptera_percent, "NA"))) %>%  
mutate(Coleoptera_percent = as.numeric(na_if(Coleoptera_percent, "NA"))) 
#Calculate average
mean_study = metadata_study  %>% 
summarise(Hymenoptera_percent = mean(Hymenoptera_percent, na.rm = T),
          Diptera_percent = mean(Diptera_percent, na.rm = T),
          Lepidoptera_percent = mean(Lepidoptera_percent, na.rm = T),
          Coleoptera_percent = mean(Coleoptera_percent, na.rm = T)) %>% 
 mutate(across(c(Hymenoptera_percent, Diptera_percent, Lepidoptera_percent, Coleoptera_percent),
                ~ ifelse(is.nan(.), NA, .))) 
#Now calculate percentage of studies that record any of these 4 orders
mean_percent_order = mean_study %>%
  summarise(
    Hymenoptera_percent_metaweb = sum(!is.na(Hymenoptera_percent)) / nrow(.) * 100,
    Diptera_percent_metaweb = sum(!is.na(Diptera_percent)) / nrow(.) * 100,
    Lepidoptera_percent_metaweb = sum(!is.na(Lepidoptera_percent)) / nrow(.) * 100,
    Coleoptera_percent_metaweb = sum(!is.na(Coleoptera_percent)) / nrow(.) * 100)

Hymenoptera_percent_metaweb = mean_percent_order$Hymenoptera_percent_metaweb
Diptera_percent_metaweb = mean_percent_order$Diptera_percent_metaweb
Lepidoptera_percent_metaweb = mean_percent_order$Lepidoptera_percent_metaweb
Coleoptera_percent_metaweb = mean_percent_order$Coleoptera_percent_metaweb

#Check percentages of hymenoptea groups
hymenoptera_percent_by_group = metadata %>% 
ungroup() %>% 
distinct(Study_id, Taxa_recorded) %>% 
summarise(Bee_studies = sum(str_detect(Taxa_recorded, "Bees")/ nrow(.) * 100),
          Bumblebee_studies = sum(str_detect(Taxa_recorded, "Bumblebees")/ nrow(.) * 100),
          Hymenoptera_studies = sum(str_detect(Taxa_recorded, "Hymenoptera")/ nrow(.) * 100))

bee_percent = hymenoptera_percent_by_group$Bee_studies
bumblee_percent = hymenoptera_percent_by_group$Bumblebee_studies
all_hymnoptera_percent = hymenoptera_percent_by_group$Hymenoptera_studies

#Check percentages of diptera groups
diptera_percent_by_group = metadata %>% 
ungroup() %>% 
distinct(Study_id, Taxa_recorded) %>% 
summarise(Syrphid_studies = sum(str_detect(Taxa_recorded, "syrphids")/ nrow(.) * 100),
          Diptera_studies = sum(str_detect(Taxa_recorded, "Diptera")/ nrow(.) * 100),
          other_diptera_studies = sum(str_detect(Taxa_recorded, "syrphids, bombylids") |
                                       str_detect(Taxa_recorded, "syrphids, conopids, tachinid flies"))/ nrow(.) * 100)

syrphid_percent = diptera_percent_by_group$Syrphid_studies
all_diptera_percent = diptera_percent_by_group$Diptera_studies
other_diptera_percent = diptera_percent_by_group$other_diptera_studies

@@

**Main Types of Variables Included:** EuPPollNet contains `r prettyNum(interaction_number, big.mark=",")` interactions between plants and pollinators from `r prettyNum(n_networks, big.mark=",")` distinct networks (i.e., distinct sampling event in space or time), which belong to `r study_number` different studies distributed across `r country_number` European countries. In addition, information about sampling methodology, habitat type, bio-climatic region, and further taxonomic rank information for both plant and pollinator species are also provided (i.e., genus, family and order).

**Spatial location and grain:** The database contains `r prettyNum(n_sites, big.mark=",")` different sampling locations from `r n_habitats` different natural and anthropogenic habitats that fall in `r n_bioregion` different bio-climatic regions. All records are geo-referenced and presented in the World Geodetic System 1984 (WGS84).

@@

The interactions between different plant and pollinator species within a community form complex networks. Macro-ecological analyses of the topology of these networks have revealed common properties, such as truncated power-law degree distributions [@jordano2003] or nestedness [@bascompte2003]. Large-scale analyses across multiple studies can quantify patterns across geographic regions [@olesen2002; @traveset2016] or environmental gradients [@ramos2010; @rech2016; @saunders2023] that cannot be examined in a single study. Although macro-ecological approaches that use ecological interactions make significant contributions to knowledge [@windsor2023], such synthesis work must consider variation across studies in the spatio-temporal nature of the data [@burkle2011; @garcia2024]. For instance, plant-pollinator studies tend to differ in sampling effort and methodology which affect the structure of the resulting plant-pollinator networks [@gibson2011; @jordano2016; @schwarz2020]. Most plant-pollinator networks have unobserved interactions [@olesen2011; @chacoff2012], and thus research that attempts to synthesize across published studies must have access to raw data on interactions in order to statistically account for sampling effort and completeness. This emphasizes the importance of providing data in its rawest possible form in datasets that will be utilized for synthesis and macro-ecological studies.

Europe is one of the continents with a larger amount of available biodiversity data [@proencca2017], yet still exhibits major gaps [@wetzel2018; @bennett2018]. While species checklists need to be treated carefully, especially at a macro-ecological scale [@grenie2023], the growing number of European plant and pollinator checklists [@reverte2023], along with occurrence data [@zattara2021], is setting a foundation for the conservation of its flora and their pollinators. However, species richness is just one component of biodiversity and documenting the interaction between plants and pollinators is essential for understanding biodiversity change [@jordano2016]. Numerous works have studied plant-pollinator interactions in the last decades, generating thousands of plant-pollinator interaction networks worldwide. Several initiatives have tried to integrate plant-pollinator interaction data into databases such as *Mangal* [@poisot2016] or *GloBI* [@poelen2014], resulting in numerous large scale comparative analyses that have enhanced our understanding of the ecology of plants and pollinators [e.g., European wild bee data trends\; @marshall2024]. Despite all these resources, Europe lacks accessible harmonized plant-pollinator interaction data that allow researchers to evaluate plant-pollinator interactions at a European level. For example, only over a dozen of European plant-pollinator networks are included in *Mangal*, while *GloBI* focuses on pairwise interactions disconnected from the community context. Assembling and curating the existing information on EU plant-pollinator networks will guide research efforts, conservation planning and will set a foundation for future global change research.

Here, we present the European Plant-Pollinator Networks database (EuPPollNet), which contains harmonized information on plant-pollinator interactions at the European level. The pollinator taxonomic groups include the main orders of entomofauna that visit and pollinate flowering plants in Europe. These comprise insect species from the orders Hymenoptera, Diptera, Lepidoptera, and Coleoptera, accounting for almost the totality of recorded interactions in EuPPollNet (`r format(round(main_orders_percent, 2), nsmall = 2)`%). To understand the scope of the database, we examined the taxonomic and sampling coverage of the different plant and pollinator species at the European level with the help of the most up-to-date species checklists and rarefaction analyses. In addition, for bees and plants, we evaluated if there is a phylogenetic signal in the presence-absence of interaction data.

EuPPollNet contains one of the largest sets of plant-pollinator networks collated to date, providing a unique opportunity to examine the prevalence of key structural metrics across networks. For example, despite the large theoretical literature on the meaning of a nested structure in plant-pollinator networks [@bascompte2007; @guimaraes2020], where specialists species interact only with subsets of generalists species, this pattern has only been empirically evaluated with a relatively small number of networks [@bascompte2003; @staniczenko2013; @payrato2019], and is still debated how structural metrics such as connectance and nestedness change across latitudes and bioclimatic regions [@olesen2002; @trojelsgaard2013; @song2017].

Overall, EuPPollNet aims to cover a wide range of taxonomic groups and habitats, while also providing other variables of interest that define the ecological context and sampling methods of the study. In addition, EuPPollNet offers a transparent and accessible workflow of its data management and species harmonization that allows the database to be reused and to expand over time. This database provides a large number of community-level networks with curated and harmonized data, distinguishing it from other currently available resources that contain plant-pollinator interactions. We expect that EuPPollNet can be used to evaluate macro-ecological processes in plant-pollinator networks, guide conservation planning, and set a baseline for global change research.

@@

The EuPPollNet database includes published and unpublished studies compiled initially by a wide number of researchers and institutions within the European continent as defined by the European Environment Agency [@stanners1995]. As this database is the result of one of the work packages of the European project Safeguard (Safeguarding European wild pollinators; https://doi.org/10.3030/101003476), first, data was directly requested from members of the Safeguard project in May 2022. Second, the request was extended to data owners outside of the project, with data collection concluding in August 2024. These other data owners were identified by direct communication with colleagues suggested by Safeguard members and by directly searching for studies on Google Scholar of under-represented regions within the database. While Google Scholar lacks reproducibility [@gusenbauer2020], it still remains the most comprehensive search engine to date [@gusenbauer2019]. This approach maximized the potential number of studies that could be incorporated in this database. The search strings used were '*plant-pollinator interactions*' and '*plant-pollinator networks*'. To maintain high quality standards that will support robust future ecological research, we only included studies that met the following criteria: 1) studies containing time- and geo-referenced records of plant-pollinator interactions; and 2) studies that quantify interactions by documenting the contact between a floral visitor - referred to as a 'pollinator' throughout the manuscript, even though pollination efficiency is not evaluated - and the reproductive structure of a specific sampled plant (i.e., phyto-centric networks).

@@

The database contains `r study_number` independent published and unpublished studies conducted during the time period `r min_year` - `r max_year` in `r country_number` different countries (**Figure 1a** and **Figure 1b**; see **Figure S1** for exact locations). The studies differ in sampling effort and methodology, and thus documenting sampling methods and sampling effort is an important feature of EuPPollNet. Most studies took place within a single flowering season (`r format(round(one_year, 2), nsmall = 2)`%), sampled a given location for an average of `r format(round(average_sampling_days, 2), nsmall=2)` days, and exclusively sampled diurnal plant-pollinator interactions, with `r  most_common_method[[1]]` being the most common sampling method (`r format(round(most_common_method[[3]], 2), nsmall = 2)`%). All the studies documented interactions with Hymenopterans (with `r format(round(all_hymnoptera_percent, 2), nsmall = 2)`% considering all Hymenopterans, `r format(round(bee_percent, 2), nsmall = 2)`% only wild bees and `r format(round(bumblee_percent, 2), nsmall = 2)`% only bumblebees), `r format(round(Diptera_percent_metaweb, 2), nsmall = 2)`% documented interactions with Dipterans (with `r format(round(all_diptera_percent, 2), nsmall = 2)`% considering all Dipterans, `r format(round(syrphid_percent, 2), nsmall = 2)`% only syrphids and `r format(round(other_diptera_percent, 2), nsmall = 2)`% recorded syrphids plus bombylids or tachinid flies), `r format(round(Lepidoptera_percent_metaweb, 2), nsmall = 2)`% with Lepidopterans, and `r format(round(Coleoptera_percent_metaweb, 2), nsmall = 2)`% with Coleopterans. The database includes a total of `r prettyNum(interaction_number, big.mark=",")` distinct interactions. Most of the pollinator species belong to the orders Hymenoptera, Diptera, Lepidoptera and Coleoptera (`r format(round(spp_percent_main_order, 2), nsmall = 2)`%), which account for nearly all interactions in this database (`r format(round(main_orders_percent, 2), nsmall = 2)`%). Species that belong to other orders (`r format(round(100-spp_percent_main_order, 2), nsmall = 2)`%) are not explored in this study as they conduct a minor fraction of the total interactions (`r format(round(100-main_orders_percent, 2), nsmall = 2)`%). Hymenoptera and Diptera contain the highest number of species comprising each approximately 1,000 species in the database. However, the majority of plant-pollinator interactions are from Hymenoptera species (`r format(round(Hymenoptera_interactions, 2), nsmall = 2)`%; **Figure 1c**). Notably, the western honey bee, _Apis mellifera_, represents `r format(round(apis_percent, 2), nsmall = 2)`% of the total interaction records from the database and an average of `r format(round(apis_average_network, 2), nsmall = 2)`% of the total interactions per network.

```{r echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, out.width="105%", fig.cap= "\\textbf{Figure 1}. (\\textbf{a}) Locations of the studies in EuPPollNet across the European continent, showing the total number of pollinator (i.e., orange heptagon) and plant (i.e., green circles) species per study. The sizes of these shapes are proportional to the respective species counts. For visualization purposes, we have selected only a single location per study. (\\textbf{b}) Number of studies by year in EuPPollNet. (\\textbf{c}) Proportion of species and interactions across the four main pollinator orders in EuPPollNet, excluding interactions from \\textit{Apis mellifera}. The total number of species and interactions is indicated in parentheses at the bottom."}

@@

p2  = ggplot(data2, aes(Year, Counts)) +
geom_col(fill = "gray25", width = 0.9, color="black", linewidth = 0.4) +
theme_light() +
theme(plot.margin=unit(c(0,0,0,0), "mm"),
  axis.title.y = element_text(face = "bold", size = 7, vjust = -2),
  axis.title.x = element_text(face = "bold", size = 8),
  axis.text.y = element_text(size = 6),
  axis.text.x = element_text(size = 6,  angle = 90), 
  strip.text.x = element_text(size=12),
   axis.ticks.x = element_line(color="black"),
  strip.background = element_rect(
  color="black", fill=NA, size=0.5, linetype="solid")) + 
ylab("Number of studies") +
coord_cartesian(expand = F, clip = "off") + 
theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  panel.background = element_blank(),
  panel.border = element_rect(colour = "black", fill=NA, size=1)) +
theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(),
panel.border = element_blank(),
axis.line = element_line(color = 'black')) +
scale_y_continuous(breaks = c(0, 4,8,12), labels = c("0", "4","8", "12"), limits = c(0,12)) +
scale_x_continuous(breaks = seq(min(data2$Year), max(data2$Year), by = 2), limits = c(min(data2$Year),max(data2$Year)))+
ggtitle("(b)") +
theme(plot.title = element_text(size = 10,hjust = -0.4, vjust = -5))

main_orders = c("Hymenoptera", "Diptera", "Lepidoptera", "Coleoptera")

#Calculate spp number and interactions separately
data1_spp = data %>% 
select(Pollinator_order, Pollinator_accepted_name, Pollinator_rank) %>% 
filter(Pollinator_order %in% main_orders) %>% 
filter(Pollinator_rank == "SPECIES") %>% 
group_by(Pollinator_order) %>% 
summarise(Species = n_distinct(Pollinator_accepted_name)/n_distinct(.))

data1_int = data %>% 
filter(!Pollinator_accepted_name == "Apis mellifera") %>% 
select(Pollinator_order, Pollinator_accepted_name, Pollinator_rank) %>% 
filter(Pollinator_order %in% main_orders) %>% 
#filter(Pollinator_rank == "SPECIES") %>% 
group_by(Pollinator_order) %>% 
summarise(Interactions = length(Pollinator_accepted_name)/ nrow(.))

data1 = left_join(data1_spp, data1_int)

#Select main orders
spp = data %>% 
select(Pollinator_order, Pollinator_accepted_name, Pollinator_rank) %>% 
filter(Pollinator_order %in% main_orders) %>% 
filter(Pollinator_rank == "SPECIES") %>% 
summarise(Species = n_distinct(Pollinator_accepted_name)) %>% 
pull()
#Select main orders
int = data %>% 
filter(!Pollinator_accepted_name == "Apis mellifera") %>% 
select(Pollinator_order, Pollinator_accepted_name, Pollinator_rank) %>% 
filter(Pollinator_order %in% main_orders) %>% 
#filter(Pollinator_rank == "SPECIES") %>% 
summarise(Interactions = length(Pollinator_accepted_name)) %>% 
pull()

#Combine Interactions and Species into a single column
data_long = pivot_longer(data1, cols = c(Interactions, Species), names_to = "Variable", values_to = "Value")

@@

#Plot
p3 = ggplot(data_long, aes(x = Variable, y = Value, fill = Pollinator_order)) + 
geom_bar(stat = "identity", width = 0.7) +
theme_classic() +
scale_fill_viridis(discrete=TRUE, direction = -1, name="Pollinator\n order") +
xlab(NULL) +
ylab("Percentage") + 
scale_y_continuous(labels = percent_format(), expand = c(0, 0)) +
theme(axis.text = element_text(size=6), 
     axis.title.y = element_text(face = "bold", size = 7, vjust = -2),  
     axis.title.x = element_text(face="bold", size = 7),
     legend.position = "top",
     legend.title = element_text(size = 5, face = "bold"),
     legend.text = element_text(size = 5),
     legend.key.size = unit(3, "mm"),
     legend.margin=margin(0,2,0,0),
     legend.box.margin=margin(-9,-9,-6,-6)) +
guides(fill=guide_legend(nrow=2,byrow=TRUE))+
ggtitle("(c)") +
theme(plot.title = element_text(size = 10,hjust = -0.4, vjust = -6)) +
scale_x_discrete(labels = c("Species\n (2,172)", "Interactions\n (358,049)"))

@@

The EuPPollNet database is available in both .csv and .rds formats and contains a total of `r nrow(my_tibble)` columns (**Table 1**), where each row represents a single interaction between a plant and a pollinator species. These columns include information about the study and network identifiers (columns 1 and 2), sampling method (3), habitat type as described by the author, and a unified habitat classification across studies (4 and 5), bioregion where the network is located (6), country, locality, and latitude-longitude coordinates (7 to 10), date of the interaction (11), number of interactions (12), taxonomic information about plants (13 to 20), taxonomic information about pollinators (21 to 28), and information about the availability of floral count data (29). The flower count data is provided in a separate file (.csv or .rds) and can be merged with the interaction data through the "Flower_data_merger" column (30). Note that although two-thirds of studies include information on floral abundance, the methods and units vary greatly across studies. To construct a plant-pollinator network matrix within a single flowering season at the site level, users should group interactions by plant and pollinator species, site, study, and year. Finally, metadata at the study level is provided in a separate file, including information about the authors, digital object identifier (if available), sampling time, and taxonomic coverage of the main pollinator groups for each study.

@@

To assess the completeness of plant and pollinator species in the EuPPollNet database at the European level, we used the aforementioned checklists for plants and pollinators. Specifically for plants, we refined the checklist to include only plants occurring in Europe and excluded taxonomic groups not associated with biotic pollination. We did this by first excluding the families considered to have exclusively a wind pollination mode [see @culley2002], and then by filtering out the genera with wind or non-biotic pollination from families that exhibit both biotic and non-biotic pollination modes. Additionally, we manually included exotic species and added unresolved species names that were not present in the accepted names of the checklist at the current version of usage. For pollinators, we only evaluated the taxonomic coverage of groups with species checklists available in Europe (i.e., bees, syrphids and butterflies). While there is not a good understanding of pollinator diversity in other taxonomic groups (e.g., beetles), it was assumed that their coverage within the database is equal to the average coverage of bees, syrphids, and butterflies (mean coverage = `r round(mean_coverage, 2)`; sd = `r round(sd_coverage,2)`). Therefore, the total number of of flower visiting species from other taxonomic groups (i.e., non-bee, non syrphid and non-butterfly flower visitors) at European level was extrapolated by assuming that their coverage is equal to the mean coverage of bees, syrphids, and butterflies. Consequently, we provide an estimate for the total number of pollinators across the European continent.

@@

To provide a general overview of the structure of plant-pollinator networks in EuPPollNet, we quantified connectance and nestedness for each network and examined how these network metrics change across different latitudes and bioclimatic regions in Europe. We selected these two network metrics as they are commonly evaluated in plant-pollinator network studies and capture structural properties with a straightforward interpretation. We implemented “standardised” versions of connectance and nestedness to account for the effect of sampling effort on network metrics. As connectance is negatively associated with network size [@jordano1987], we evaluated how network connectance was associated with the number of species (i.e., log of geometric mean of plants and pollinators) and extracted the residuals from this association (i.e., residual connectance) as a measurement of corrected connectance. The relationship between residual connectance and species richness was investigated using a beta regression. This approach was chosen because connectance displays a non-normal distribution with continuous values bounded between 0 and 1. The model was implemented using the betareg package [@cribari2010]. We used NODFc to compare nestedness across networks, as it corrects by connectance and the number of species in comparison to other nestedness metrics that change with network size [@song2017]. This metric was calculated using the **maxnodf** package [@hoeppke2021]. Both residual connectance and NODFc were used as dependent variables to evaluate their association with latitude. In addition, to quantify how connectance and nestedness change with network size, we determined their association with the number of species per network using the Kendall rank correlation coefficient.
